```{r loadData, include=FALSE}
options(stringsAsFactors=FALSE)
require(synapseClient)
require(rGithubClient)
require(Biobase)
require(GEOquery)
require(ggplot2)

## GET THE LOCATION OF THIS FILE ON GITHUB
crcscRepo <- getRepo("/Sage-Bionetworks/crcsc")
rmdUrl <- getPermlink(crcscRepo, "groups/G/dataQc/frenchClinical.Rmd")

synFolder <- "syn2019116"

gseAll <- getGEO("GSE39582", GSEMatrix=TRUE)
gseSet <- gseAll[[1]]
# gseSet <- combine(gseAll[[1]], gseAll[[2]])
# gseSet <- combine(gseSet, gseAll[[3]])

clin <- pData(gseSet)

## PARSE THE CLINICAL DATA
newClin <- data.frame(row.names=rownames(clin))
theseCols <- grep("characteristic", names(clin), value=TRUE)
p <- lapply(as.list(theseCols), function(x){
  strsplit(as.character(clin[, x]), ": ", fixed=T)
})
for( i in 1:length(p) ){
  thisVar <- unique(sapply(p[[i]], "[[", 1))
  newClin[, thisVar] <- sapply(p[[i]], "[[", 2)
}

clinOut <- data.frame(id=rownames(newClin),
                      dataset=newClin$dataset,
                      age=floor(as.numeric(newClin$age.at.diagnosis)),
                      gender=newClin$Sex,
                      stage=newClin$tnm.stage,
                      tStage=NA,
                      nStage=NA,
                      mStage=NA,
                      tumorLocation=newClin$tumor.location,
                      rfsMo=newClin$rfs.delay/365*12,
                      rfsStat=newClin$rfs.event,
                      batch=newClin$normalizationcombatbatch,
                      microsatelite=newClin$mmr.status,
                      cimp=NA,
                      adjChemo=newClin$chemotherapy.adjuvant,
                      krasStatus=newClin$kras.mutation.status,
                      krasProtein=newClin$kras.mutation.protein,
                      brafStatus=newClin$braf.mutation.status,
                      brafProtein=newClin$braf.mutation.protein,
                      tp53Status=newClin$tp53.mutation.status,
                      tp53Protein=newClin$tp53.mutation.protein)

## WRITE OUT AN ACTIVITY THAT CAPTURES WHAT WAS USED IN OUR ANALYSIS
act <- Activity(name="Clinical curation knitr script", used=list(list(url=rmdUrl, name=basename(rmdUrl), wasExecuted=TRUE)))
act <- synStore(act)

## CLINICAL FILE
fClinFile <- file.path(tempdir(), "FRENCH_clinical.tsv")
write.table(clinOut, file=fClinFile, sep="\t", quote=FALSE, row.names=FALSE)

clinFile <- File(path=fClinFile, parentId=synFolder)
generatedBy(clinFile) <- act
clinFile <- synStore(clinFile)
```
##! FRENCH Clinical Data

This wiki summarizes the clinical data for patients which have expression data available and QCed in this project. The resulting tab delimited text file [`r clinFile@properties$name`](#!Synapse:`r clinFile@properties$id`) contains clinical information on these `r nrow(clinOut)` samples.

${toc}


## Dataset
```{r dataset, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=dataset)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
## Age Distribution
```{r age, echo=FALSE, fig.height=5, fig.width=12}
ggplot(data=clinOut) + geom_density(aes(x=age), fill="blue", alpha=0.5)
```
## Gender
```{r gender, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=gender)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
## Tumor Stage
```{r stage, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=stage)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
## Tumor Location
```{r location, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=tumorLocation)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
## MSI
```{r msi, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=microsatelite)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
## Adjuvent Chemotherapy
```{r chemo, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=adjChemo)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
## KRAS
####! Mutation Status
```{r krasStatus, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=krasStatus)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
####! Protein
```{r krasProtein, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=krasProtein)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
## BRAF
####! Mutation Status
```{r brafStatus, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=brafStatus)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
####! Protein
```{r brafProtein, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=brafProtein)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
## tp53
####! Mutation Status
```{r tp53Status, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=tp53Status)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
####! Protein
######! too many different protein locations to graphically display
```{r tp53Protein, include=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=tp53Protein)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```

.