```{r loadData, include=FALSE}
## FUNCTIONS TO EXTRACT DATA OBJECTS FROM SYNAPSE AND CURATE CLINICAL INFO
#####
## ANALYST: BRIAN M. BOT
#####
options(stringsAsFactors=FALSE)
require(synapseClient)
require(rGithubClient)
require(ggplot2)

## GET THE LOCATION OF THIS FILE ON GITHUB
crcscRepo <- getRepo("/Sage-Bionetworks/crcsc")
rmdUrl <- getPermlink(crcscRepo, "groups/G/dataQc/kfsysccClinical.Rmd")

## SYNAPSE FOLDER
synFolder <- "syn2171230"

kfExprSyn <- synGet("syn2169565")
kfExprHead <- read.delim(kfExprSyn@filePath, header=F, as.is=TRUE, nrows=1)

thesePatients <- as.character(kfExprHead[1, -1])

## GRAB THE CLINICAL DATA
## READ IN THE CLINICAL DATA
clinSyn <- synGet("syn2025168")
clin <- read.delim(getFileLocation(clinSyn), as.is=T, na.strings=c(" ", "NA", ""))
rownames(clin) <- paste(clin$SN, "CEL", sep=".")

## SUBSET TO COHORT OF INTEREST
clin <- clin[thesePatients, ]
clin$osStat <- ifelse(clin$Death == 0 | is.na(clin$Death), 0, 1)
clin$osMo <- clin$FT*12
# clin$dfsStat <- ifelse(clin$Relap_M == 0 | is.na(clin$Relap_M), 0, 1)
# clin$dfsMo <- clin$MET
# clin$dfsStat[ clin$dfsStat==0 & clin$osStat==1 ] <- 1
# clin$dfsMo[ clin$dfsStat==0 & clin$osStat==1 ] <- clin$osMo[ clin$dfsStat==0 & clin$osStat==1 ]


clinOut <- data.frame(id=rownames(clin),
                      age=clin$AGE,
                      gender=NA,
                      stage=gsub(" ", "", clin$Stage),
                      tStage=as.character(clin$StageT),
                      mStage=as.character(clin$StageM),
                      nStage=as.character(clin$StageN),
                      tumorLocation=ifelse(clin$Site==1, "Ascending", ifelse(clin$Site==2, "Transverse", ifelse(clin$Site==3, "Descending", ifelse(clin$Site==4, "Recto-sigmoid", NA)))),
                      dfsMo=NA,
                      dfsStat=NA,
                      osMo=clin$osMo,
                      osStat=clin$osStat,
                      batch=NA,
                      microsatelite=NA,
                      cimp=NA,
                      adjChemo=NA,
                      krasCodon12=clin$KRAS_Codon12,
                      krasCodon13=clin$KRAS_Codon13)

## WRITE OUT AN ACTIVITY THAT CAPTURES WHAT WAS USED IN OUR ANALYSIS
act <- Activity(name="Clinical curation knitr script", used=list(kfExprSyn, clinSyn, list(url=rmdUrl, name=basename(rmdUrl), wasExecuted=TRUE)))
act <- synStore(act)

## CLINICAL FILE
kfClinFile <- file.path(tempdir(), "KFSYSCC_clinical.tsv")
write.table(clinOut, file=kfClinFile, sep="\t", quote=FALSE, row.names=FALSE)

clinFile <- File(path=kfClinFile, parentId=synFolder)
generatedBy(clinFile) <- act
clinFile <- synStore(clinFile)
```
##! KFSYSCC Clinical Data

This wiki summarizes the clinical data for patients which have expression data available and QCed in this project. The resulting tab delimited text file [`r clinFile@properties$name`](#!Synapse:`r clinFile@properties$id`) contains clinical information on these `r nrow(clinOut)` samples.

${toc}


## Age Distribution
```{r age, echo=FALSE, fig.height=5, fig.width=8}
ggplot(data=clinOut) + geom_density(aes(x=age), fill="blue", alpha=0.5)
```
## Tumor Stage
####! Overall
```{r stage, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=stage)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
####! T Stage
```{r tstage, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=tStage)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
####! N Stage
```{r nstage, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=nStage)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
####! M Stage
```{r mstage, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=mStage)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
## Tumor Location
```{r location, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=tumorLocation)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
## KRAS
####! Codon 12
```{r krasCodon12, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=krasCodon12)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```
####! Codon 13
```{r krasCodon13, echo=FALSE, fig.height=5, fig.width=6}
ggplot(data=clinOut, aes(x=factor(1), fill=krasCodon13)) + 
  geom_bar(width=1) + 
  coord_polar(theta="y") +
  xlab('') + ylab('')
```

.