```{r loadData, include=FALSE}
## SCRIPT TO QC AND PROCESS THE KFSYSCC DATASET
#####
## ANALYST: BRIAN M. BOT
#####
require(synapseClient)
require(rGithubClient)
require(affy)
require(affyPLM)
require(limma)
require(chron)
require(corpcor)
require(ggplot2)
require(hgu133plus2.db)

## GET THE LOCATION OF THIS FILE ON GITHUB
crcscRepo <- getRepo("/Sage-Bionetworks/crcsc")
rmdUrl <- getPermlink(crcscRepo, "dataQc/kfsysccExpression.Rmd")

## CONVENIENCE FUNCTION FOR SVD EVALUATIONS
fs <- function(x){
  require(corpcor)
  u <- fast.svd(t(scale(t(x), scale = FALSE)), tol = 0)
  u$d <- u$d^2/sum(u$d^2)
  return(u)
}

synFolder <- "syn2025137"

## GRAB THE TAR FILE - UNTAR IN TEMP DIRECTORY
kfEnt <- synGet("syn2025141")
kfDir <- tempfile(pattern="kfDir")
dir.create(kfDir)
untar(getFileLocation(kfEnt), exdir = kfDir)
theseFiles <- list.celfiles(kfDir, recursive=T, full.names=T)

## READ IN CEL FILES
kfExpr <- ReadAffy(filenames=theseFiles)

scanDate <- protocolData(kfExpr)$ScanDate
scanDate <- chron(gsub(" .*", "", scanDate))
pData(kfExpr) <- data.frame(scanDate=scanDate, row.names=sampleNames(kfExpr))


## ORDER SAMPLES IN CHRONOLOGICAL ORDER OF SCAN DATE
kfRaw <- kfExpr[, order(pData(kfExpr)$scanDate)]

## PERFORM SOME PRELIMINARY QC
pset <- fitPLM(kfRaw)
```
## Initial assessment of expression data
These samples were QCed initially using the `affyPLM` package. The following plot shows the Relative Log Expression (RLE) and Normalized Unscaled Standard Errors (NUSE) plots for all samples. Given the high NUSE for samples processed before 01/01/2007, those `r sum(pData(kfRaw)$scanDate <= "01/01/07")` samples were removed from all further analyses.
```{r plmPlots, echo=FALSE, fig.height=10, fig.width=6}
par(mfrow=c(2,1))
RLE(pset, main="RLE for CRC322")
NUSE(pset, main="NUSE for CRC322")
rm(pset)
```
```{r subset, include=FALSE}
# REMOVE EARLY SAMPLES; LARGE SE FROM NUSE PLOT
kfRaw <- kfRaw[ , pData(kfRaw)$scanDate > "01/01/07" ]
exprSet <- rma(kfRaw)
expr <- exprs(exprSet)

## SVD ON EXPRESSION MATRIX -- ASSESS OVERALL STRUCTURE AND POSSIBLE LATENT STRUCTURE
s <- fs(expr)
```
## First pass normalization
We then ran `rma` on the resulting `r ncol(expr)` samples which consolidates the feature space down to `r nrow(expr)` probesets. Singular Value Decomposition (SVD) was performed on the entire resulting expression matrix to assess its structure.

```{r rawSvd, echo=FALSE, fig.height=5, fig.width=7}
qplot(1:length(s$d), s$d,
      xlab="eigen gene",
      ylab="% variance explained")
qplot(s$v[, 1], s$v[, 2],
      xlab="1st svd",
      ylab="2nd svd")
```
```{r saveObject, include=FALSE}
## WRITE OUT AN ACTIVITY THAT CAPTURES WHAT WAS USED IN OUR ANALYSIS
act <- Activity(name="QC knitr script", used=list(kfEnt, list(url=rmdUrl, name=basename(rmdUrl), wasExecuted=TRUE)))
act <- synStore(act)

## EXPRESSION FILE
expr <- as.data.frame(expr)
tmpNames <- colnames(expr)
expr$feature <- rownames(expr)
expr <- expr[, c("feature", tmpNames)]
kfExprFile <- file.path(tempdir(), "KFSYSCC_expression.tsv")
write.table(expr, file=kfExprFile, sep="\t", quote=FALSE, row.names=FALSE)

exprFile <- File(path=kfExprFile, parentId=synFolder)
generatedBy(exprFile) <- act
exprFile <- synStore(exprFile)
```
## QC output
The resulting tab delimited text files contain expression values on `r nrow(expr)` genes and `r ncol(expr)-1` samples.


.