##! TCGA CRC Clinical Data

This wiki summarizes the clinical data for patients which have RNA-seq data available and QCed in this project.
${toc}
```{r loadData, include=FALSE}
## FUNCTIONS TO EXTRACT DATA OBJECTS FROM SYNAPSE AND CURATE CLINICAL INFO
#####
## ANALYST: BRIAN M. BOT
#####
options(stringsAsFactors=FALSE)
require(synapseClient)
require(rGithubClient)
require(ggplot2)

## GET THE LOCATION OF THIS FILE ON GITHUB
crcscRepo <- getRepo("/Sage-Bionetworks/crcsc")
rmdUrl <- getPermlink(crcscRepo, "dataQc/tcgaCrcClinical.Rmd")

## READ IN FILES AS FORMATED BY TCGA PAN CANCER GROUP - UTILIZING OLD 'DATA' OBJECTS INSTEAD OF FILES
loadTCGAFile <- function(ent){
  df <- read.delim(file.path(ent$cacheDir, ent$files), header=F, as.is=T, na.strings=c("", "NA", " "))
  colnames(df) <- as.character(df[1, ])
  df <- df[-1, ]
  rownames(df) <- as.character(df[, 1])
  df <- df[, -1]
  return(df)
}


## SYNAPSE FOLDER FOR THE TCGA DATA
synFolder <- "syn2023932"

crcRNAseqSyn <- synGet("syn2161141")
crcRNAseqHead <- read.delim(crcRNAseqSyn@filePath, header=F, as.is=TRUE, nrows=1)

thesePatients <- as.character(crcRNAseqHead[1, -1])

## GRAB THE CLINICAL DATA
coadClinSyn <- synGet("syn1446080")
coadClin <- loadTCGAFile(coadClinSyn)
readClinSyn <- synGet("syn1446153")
readClin <- loadTCGAFile(readClinSyn)
coadClin$rns <- rownames(coadClin)
readClin$rns <- rownames(readClin)

crcClin <- merge(x=coadClin, y=readClin, all=T)
rownames(crcClin) <- crcClin$rns
crcClin$rns <- NULL
clin <- crcClin

## SUBSET TO COHORT OF INTEREST
clin <- clin[thesePatients, ]

clinOut <- data.frame(id=clin$bcr_patient_barcode,
                      age=as.numeric(clin$age_at_initial_pathologic_diagnosis),
                      gender=tolower(clin$gender),
                      stage=clin$tumor_stage,
                      tStage=clin$primary_tumor_pathologic_spread,
                      nStage=clin$lymphnode_pathologic_spread,
                      mStage=clin$distant_metastasis_pathologic_spread,
                      tumorLocation=clin$anatomic_site_colorectal,
                      dfsMo=NA,
                      dfsStat=NA,
                      osMo=ifelse(clin$vital_status=="DECEASED",
                                  as.numeric(clin$days_to_death[clin$vital_status == "DECEASED"])*365/12,
                                  as.numeric(clin$days_to_last_known_alive[clin$vital_status != "DECEASED"])*365/12),
                      osStat=ifelse(clin$vital_status=="DECEASED", 1, 0),
                      batch=NA,
                      microsatelite=clin$mononucleotide_and_dinucleotide_marker_panel_analysis_status,
                      cimp=NA,
                      adjChemo=NA,
                      krasStatus = clin$"KRAS:mutation",
                      brafStatus = clin$"BRAF:mutation",
                      alkStatus = clin$"ALK:mutation",
                      egfrStatus = clin$"EGFR:mutation",
                      nrasStatus = clin$"NRAS:mutation",
                      erbb2Status = clin$"ERBB2:mutation")
```
### Age Distribution
```{r age, echo=FALSE, fig.height=4, fig.width=6}
ggplot(data=clinOut) + geom_density(aes(x=age), fill="blue", alpha=0.5)
```
### Gender
```{r gender, echo=FALSE, results='asis'}
synTable(table(clinOut$gender))
```
### Tumor Stage
```{r stage, echo=FALSE, results='asis'}
synTable(table(clinOut$stage))
synTable(table(clinOut$tStage))
synTable(table(clinOut$nStage))
synTable(table(clinOut$mStage))
```
### Tumor Location
```{r location, echo=FALSE, results='asis'}
synTable(table(clinOut$tumorLocation))
```
### MSI
```{r msi, echo=FALSE, results='asis'}
synTable(table(clinOut$microsatelite))
```
### Mutation Status
#### KRAS
```{r kras, echo=FALSE, results='asis'}
synTable(table(clinOut$krasStatus))
```
#### BRAF
```{r braf, echo=FALSE, results='asis'}
synTable(table(clinOut$brafStatus))
```
#### NRAS
```{r nras, echo=FALSE, results='asis'}
synTable(table(clinOut$nrasStatus))
```
#### EGFR
```{r egfr, echo=FALSE, results='asis'}
synTable(table(clinOut$egfrStatus))
```
#### ALK
```{r alk, echo=FALSE, results='asis'}
synTable(table(clinOut$alkStatus))
```
#### ERBB2
```{r erbb2, echo=FALSE, results='asis'}
synTable(table(clinOut$erbb2Status))
```
```{r saveData, include=FALSE}
## WRITE OUT AN ACTIVITY THAT CAPTURES WHAT WAS USED IN OUR ANALYSIS
act <- Activity(name="Clinical curation knitr script", used=list(crcRNAseqSyn, readClinSyn, coadClinSyn, list(url=rmdUrl, name=basename(rmdUrl), wasExecuted=TRUE)))
act <- synStore(act)

## CLINICAL FILE
tcgaCrcClinFile <- file.path(tempdir(), "TCGACRC_clinical.tsv")
write.table(clinOut, file=tcgaCrcClinFile, sep="\t", quote=FALSE, row.names=FALSE)

clinFile <- File(path=tcgaCrcClinFile, parentId=synFolder)
generatedBy(clinFile) <- act
clinFile <- synStore(clinFile)
```
### Clinical Data File
This resulting tab delimited text file [`r clinFile@properties$name`](#!Synapse:`r clinFile@properties$id`) contains clinical information on these `r nrow(clinOut)` samples.


.